---
title: "Graphing Six Nations History"
date: "2024-03-18"
draft: true
execute: 
  eval: true
---


```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(readr)

sixn <- tibble(
  year = 2000:2024,
  england = c(
    1, 1, 2, 1, 3, 4, 4, 3, 2, 2, 3, 1, rep(2, 4), 1, 1, 5, 2, 1, 5, 3, 4, 3
    ),
  france = c(
    2, 5, 1, 3, 1, 2, 1, 1, 3, 3, 1, 2, 4, 6, 4, 4, 5, 3, 4, 4, 2, 2, 1, 2, 2
    ),
  ireland = c(
    3, 2, 3, 2, 2, 3, 2, 2, 4, 1, 2, 3, 3, 5, 1, 1, 3, 2, 1, 3, 3, 3, 2, 1, 1
    ),
  italy = c(rep(6, 3), 5, 5, 6, 6, 4, rep(6, 4), 5, 4, 6, 5, rep(6, 8), 5),
  scotland = c(
    5, 3, 4, 4, 6, 5, 3, 6, rep(5, 4), 6, 3, 5, 6, 5, 5, 4, 5, rep(4, 3), 3, 4
    ),
  wales = c(
    4, 4, 5, 6, 4, 1, 5, 5, 1, rep(4, 3), 1, 1, 3, 3, 2, 5, 2, 1, 5, 1, 5, 5, 6
    )
)
```

```{r}
head(sixn)
```

```{r}
sixn_long <- pivot_longer(sixn, cols = c(england:wales))
```

```{r}
head(sixn_long)
```
```{r}
sixn_long <- sixn_long |> mutate(value = as.character(value))
```

```{r}
ggplot(sixn_long, aes(x = year, y = value, group = name, colour = name)) +
  geom_line() +
  geom_point()
```

```{r}
ggplot(sixn_long, aes(x = year, y = value, group = name, colour = name)) +
  geom_line() +
  geom_point() +
  scale_y_discrete(limits=rev)
```

```{r}
sixn_long <- mutate(
  sixn_long, year = as.character(year)
)
```


```{r}
ggplot(sixn_long, aes(x = year, y = value, group = name, colour = name)) +
  geom_line() +
  geom_point() +
  scale_y_discrete(limits=rev)
```

```{r message=FALSE}
library(lubridate)
library(glue)

sixn_dates <- sixn_long |> 
  mutate(
    year = glue("{year}-01-01"),
    year = as_datetime(year)
  )
```

```{r}
ggplot(sixn_dates, aes(x = year, y = value, group = name, colour = name)) +
  geom_line() +
  geom_point() +
  scale_y_discrete(limits=rev) +
  scale_x_datetime(date_breaks = "2 years", date_labels = "%Y")
```
Scotland's navy is #1D4976, Ireland's crest #296044, the famous Wales red is #DB4733, we'll go with white for England, which is #FFFFFF.

Both France and Italy use a few colours, Italy's blue from their crest is #275DA1 and France's blue is #133472. That's three blues in there, let's hope we can tell them apart. 

```{r}
ggplot(sixn_dates, aes(x = year, y = value, group = name, colour = name)) +
  geom_line() +
  geom_point() +
  scale_y_discrete(limits=rev) +
  scale_x_datetime(date_breaks = "2 years", date_labels = "%Y") +
  scale_colour_manual(values = c("#FFFFFF", "#133472", "#296044", "#275DA1", "#1D4976", "#DB4733"))
```


```{r}
ggplot(sixn_dates, aes(x = year, y = value, group = name, colour = name)) +
  geom_line() +
  geom_point() +
  scale_y_discrete(limits=rev) +
  scale_x_datetime(date_breaks = "2 years", date_labels = "%Y") +
  scale_colour_manual(values = c("#FFFFFF", "#133472", "#296044", "#275DA1", "#1D4976", "#DB4733"))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```
```{r}
ggplot(sixn_dates, aes(x = year, y = value, group = name, colour = name)) +
  geom_line() +
  geom_point() +
  scale_y_discrete(limits=rev) +
  scale_x_datetime(date_breaks = "2 years", date_labels = "%Y") +
  scale_colour_manual(values = c("#FFFFFF", "#133472", "#296044", "#275DA1", "#1D4976", "#DB4733")) +
  theme(
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#d9d9d9", colour = "#8c8c8c")
    )
```

```{r message=FALSE, warning=FALSE}
sixn_dates |> 
  mutate(
    name = str_to_title(name),
    yy = year(year),
    slam = case_when(
      yy %in% c(2002, 2004, 2010, 2022) & name == "France" ~ TRUE,
      yy %in% c(2003, 2016) & name == "England" ~ TRUE,
      yy %in% c(2005, 2008, 2012, 2019) & name == "Wales" ~ TRUE,
      yy %in% c(2009, 2018, 2023) & name == "Ireland" ~ TRUE,
      .default = FALSE
      )
    ) |> 
  ggplot(aes(x = year, y = value, group = name, fill = slam)) +
  geom_line() +
  geom_point(shape = 21, size = 3) +
  scale_y_discrete(limits=rev) +
  scale_x_datetime(date_breaks = "4 years", date_labels = "%Y") +
  facet_wrap(~name, axis.labels = "all_y") +
  scale_fill_manual(values = c("ghostwhite", "gold")) +
  guides(fill = "none") +
  geom_smooth(se = FALSE) +
  theme_minimal() +
  labs(x = NULL, y = NULL)
```



## ToL article

So Barnes talks about point difference (https://www.thetimes.co.uk/article/ireland-defence-coach-simon-easterby-first-pick-lions-tour-zjf26pkn0). Let's get that from data copied from https://raw.githubusercontent.com/pymc-devs/pymc-examples/main/examples/data/rugby.csv.

So that's nice but it's only a few years. Let's go to Wikipedia (support it!) to get all the years. 
```{r}
library(rvest)
library(glue)
library(purrr)

period <- 2000:2024

urls <- map(period, ~{
  glue("https://en.wikipedia.org/wiki/{.x}_Six_Nations_Championship")
})

pages <- map(urls, read_html)
```

```{r warning=FALSE, message=FALSE}
tbl <- '//*[@id="mw-content-text"]/div[1]/h2[3]/following-sibling::table'

results <- map(pages, ~{html_node(.x, xpath=tbl)}) |> 
  map(html_table)

for (i in seq_along(period)) {
  results[[i]]$year <- period[i]
}

rugger <- list_rbind(results) |> 
  select(Pos:year, -T) |> 
  mutate(
    PD = PF - PA,
    slam = case_when(
      year %in% c(2002, 2004, 2010, 2022) & Team == "France" ~ 0.25,
      year %in% c(2003, 2016) & Team == "England" ~ 0.25,
      year %in% c(2005, 2008, 2012, 2019) & Team == "Wales" ~ 0.25,
      year %in% c(2009, 2018, 2023) & Team == "Ireland" ~ 0.25,
      .default = 0
      )
    )
```

```{r}
head(rugger)
```

You can get it [here](https://raw.githubusercontent.com/RobertMyles/blogdata/master/sixnations.csv). 

```{r}
rugger |> 
  ggplot(aes(x = PD, group = Team, fill = Team)) +
  geom_boxplot() +
  theme_minimal()
```

ok, so let's create a "dominance index". We'll go naively here since we don't want to go too deep down the rabbit-hole of creating indices. I think a Grand Slam is generally regarding as a major achievement, so let's use that, as well as the points difference (PD), which we can rescale. Since `slam` is 0.25 or 0, let's rescale `PD` to between 0 and 1. We'll add the two to get our index. 

```{r message=FALSE, warning=FALSE}
library(scales)

rugger <- rugger |> mutate(
  PD = rescale(PD, to = c(0, 1)),
  dominance = PD + slam
)
```

```{r}

rugger |> 
  arrange(desc(dominance)) |> 
  select(Team, year, dominance) |> 
  head()
```

This dominance value is what we'll model with a time-series.  

```{r}
rugger |> 
  ggplot(aes(x = year, y = dominance, group = Team, colour = Team)) +
  geom_line() +
  theme_minimal()
```
A potential model could be:

$$y = y_{t-1} + \alpha + \text{slam} + \text{points difference}$$ 


