---
title: "tidyRSS v2.0.0"
author: "Robert McDonnell"
date: "2020-03-07"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyRSS)
```

[tidyRSS](https://github.com/RobertMyles/tidyRSS) was my first R package published to CRAN, successfully submitted some 3 years ago. In hindsight, there were so many things I did wrong that it's just kind of hilarious. But that's how you learn, and I have learned a _lot_ from tidyRSS. I've made quite a few other R packages since and none have come close in terms of the hassle and stress of tidyRSS, which is partly down to XML feeds themselves -- wild, untamed things.  

I've been meaning to upgrade to a new version for some time now, to clean up the code, introduce better testing and to separate out the [Simple Features](https://github.com/r-spatial/sf/) functionality into another package. (It's [here](https://github.com/RobertMyles/tidygeoRSS), have a look). It was going to be version 1.3.0, but there are so many little changes and one big breaking change, so I've gone for version 2.0.0. ([Why the numbers??](https://semver.org/)) 

I probably should have soft-deprecated the functionality I'm removing, but since it took me so long to get a space in my schedule to do this, I decided to just do it all at once. Sorry if it broke anyone's code, sincerely. But I think this is a positive change, and with tidygeoRSS, all the previous functionality is there, just present in two packages and not one.  

So what's tidyRSS for? Well, if you'd like to use RSS, Atom & JSON feeds as data, tidyRSS is for thee. This data is returned as a [_tibble_](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html) and are as 'tidy' as possible, meaning that they are easy to use downstream in your work. Depending on the feed itself, this can be a rich source of interesting data, particularly the geo-feeds. [Here's](https://towardsdatascience.com/collecting-news-articles-through-rss-atom-feeds-using-python-7d9a65b06f70) an example of collecting news with Python from RSS feeds, for example. And [here's](https://rss.com/blog/popular-rss-feeds/) a list of popular feeds from rss.com. Let's take one of these as an example to work with.  

You can install tidyRSS with: 

```{r eval = FALSE}
install.packages("tidyRSS")
```

And load it with `library(tidyRSS)`. Now let's have alook at one of the feeds from rss.com. I've chosen the New York Times, specifically the Americas section. The feed url is: [https://rss.nytimes.com/services/xml/rss/nyt/Americas.xml](https://rss.nytimes.com/services/xml/rss/nyt/Americas.xml). We just need to pass this url to `tidyfeed()`, the main function in tidyRSS:  

```{r}
nyt <- tidyfeed("https://rss.nytimes.com/services/xml/rss/nyt/Americas.xml")
```

